---
title: "Creating LDF Resources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating LDF Resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes how to create ldf resources from SPARQL queries or RDF files. You might like to first read the [Working with LDF Resources](working-with-ldf-resources.html) vignette to understand how RDF resources are represented in LDF.

```{r setup}
library(ldf)
```

## Downloading Resources with SPARQL

You can create resources by downloading a table of descriptions with a SPARQL `SELECT` query.

As an example, lets download some music genres from [dbpedia](http://dbpedia.org). This query will find 100 things, identified by their `uri` that are music genres, along with their `label` and a descriptive `comment`. We'll look for the English version of the latter two strings.

```{r}
music_genres_query <- "
PREFIX : <http://dbpedia.org/ontology/>

SELECT *
WHERE {
  ?uri a :MusicGenre;
    rdfs:label ?label;
    rdfs:comment ?comment
    .
  FILTER langMatches(lang(?label), 'EN')
  FILTER langMatches(lang(?comment), 'EN')
} LIMIT 100
"
```

We can use the `query()` function to execute the query and parse the results:

```{r eval=F}
music_genre_results <- query(music_genres_query, endpoint="http://dbpedia.org/sparql/")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(require(vcr, quietly = TRUE))
invisible(vcr::use_cassette("music-genres", {
  music_genre_results <- query(music_genres_query, endpoint="http://dbpedia.org/sparql/")
}))
```

This is what the first few results look like:

```{r}
head(music_genre_results)
```

We can then create resources for these:

```{r}
music_genres <- resource(music_genre_results$uri, data=music_genre_results)
```

Which we can then manipulate within R:

```{r}
# find music genres where the description mentions "dance"
music_genres[grep("dance", property(music_genres, "comment"))]
```

## Reading RDF with rdflib

We can create resources from serialised RDF files too.

To read RDF into R we can use the [rdflib](https://cran.r-project.org/web/packages/rdflib/) package. This in turn uses the [redland](https://cran.r-project.org/web/packages/redland/) package to provide bindings to the C library of the same name, and the [jsonld](https://cran.r-project.org/web/packages/jsonld/) package for JSON-LD serialisations.

Let's load up an example from that package.

```{r}
library(rdflib)

article_rdf <- rdf_parse(system.file("extdata", "ex.xml", package="rdflib", mustWork=TRUE))
```

This creates a list containing pointers to a redland world and model objects. We can take a peak at the statements with `rdflib::print.rdf()` (this serialises the data again and prints back the result):

```{r}
print(article_rdf, format="turtle")
```

The contents is too big to display here, but you can see from the [rdf file itself](https://github.com/ropensci/rdflib/blob/master/inst/extdata/ex.xml), that the data describes a journal article: https://doi.org/10.1002/ece3.2314.

The description is a graph, not a table. It's not a tidy collection of similarly shaped objects. We've got the article itself, and nested descriptions of related resources.

We can identify the different resource types with a query:
```{r}
rdf_query(article_rdf, "SELECT * WHERE { ?s a ?type }")
```

Here we can see the description also includes the journal in which the article is published and the creators.

We could gather all of these entities into a single resource vector, but the descriptions wouldn't overlap. The creators don't have `prism:issn` identifiers and the journal doesn't have a `foaf:familyName`.

Instead it makes more sense to split these entities into separate vectors. We'll focus on the creators, since there are several of them. We can do this with a query:

```{r}
creators_triples <- rdf_query(article_rdf, "SELECT * WHERE { ?s a <http://xmlns.com/foaf/0.1/Person>; ?p ?o }")
```

Now we have a table of statements about the creators:

```{r}
creators_triples
```

We can tabulate these statements into a tidy data frame with one row per creator, and one column per property.

```{r}
library(tidyr)

(creators_description <- creators_triples %>% 
  spread("p","o"))
```

We could proceed using the full URIs as properties, but it's nicer to replace these with shorter strings that don't need escaping with backticks:

```{r}
library(dplyr)

(creators_description <- creators_description %>% 
  rename(uri=s,
         type=`http://www.w3.org/1999/02/22-rdf-syntax-ns#type`,
         family_name=`http://xmlns.com/foaf/0.1/familyName`,
         given_name=`http://xmlns.com/foaf/0.1/givenName`,
         name=`http://xmlns.com/foaf/0.1/name`))
```

We could have done the same transformation within the select query:

```{r}
describe_creator = "
PREFIX foaf: <http://xmlns.com/foaf/0.1/>
SELECT * WHERE { 
  ?uri a <http://xmlns.com/foaf/0.1/Person>;
    a ?type;
    foaf:familyName ?family_name;
    foaf:givenName ?given_name;
    foaf:name ?name;
    .
}
"

(creators_description <- rdf_query(article_rdf, describe_creator))
```

We can then use this to create the resource vector:

```{r}
(creators <- resource(creators_description$uri, creators_description))
```
