---
title: "Introduction to LDF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to LDF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ldf)
```

Linked-data uses the [Resource Description Framework](https://www.w3.org/TR/rdf11-concepts/) (RDF) to identify resources with [Uniform Resource Identifiers](https://en.wikipedia.org/wiki/Uniform_Resource_Identifier) (URIs) and describe them with a set of statements, each specifying the value of a given property for the resource. We can represent this in R using a character vector for the URIs together with a data frame for the descriptions. That data frame should include a URI column to identify the resource being described in each row.

```{r}
uris <- c("http://example.net/id/apple",
          "http://example.net/id/banana",
          "http://example.net/id/carrot")
labels <- c("Apple","Banana","Carrot")
descriptions <- data.frame(uri=uris, label=labels)

food <- resource(uris, descriptions)
```

The `resource()` constructor returns a `ldf_resource` object that has a variety of methods defined on it, including the `format()` generic which allows us to use the labels instead of the URIs when printing to the console.

```{r}
(picnic <- data.frame(food=food, quantity=c(3,2,0)))
```

The contents of the vector itself can vary from the attached descriptions this allows you to repeat values without needing to duplicate descriptions:

```{r}
(kitchen <- data.frame(
  dish=c("Fruit Salad", "Fruit Salad",
         "Carrot Salad", "Carrot Salad"),
  food=resource(c("http://example.net/id/apple", "http://example.net/id/banana",
                  "http://example.net/id/apple", "http://example.net/id/carrot"),
                descriptions),
  quantity=c(2,2,1,3)))
```

The underlying identity of each resource in the vector can be retrieved with `uri()`:

```{r}
uri(food)
```

There's also the `curie()` function for retreiving URIs compacted with prefixes (see also: `default_prefixes()`):

```{r}
curie(food, prefixes=c(food="http://example.net/id/"))
```

You can retrieve the descriptions with `description()`:

```{r}
description(food)
```

In order to access individual properties from the resource's description, you can use `property()`:

```{r}
property(food, "label")
```

The second argument is the name of a column from the description.

Since label is such a commonly used property, there's also a function provided for it: `label()`. You can use these functions to perform operations on resources in terms of their descriptions:

```{r}
food[label(food) == "Apple"]
```

We use the label to pretty print linked data frames with `format.ldf_resource()`:

```{r}
format(kitchen)
```

Because the base type of resources is character, R will tend to dispatch on this basis. To prevent functions from base R (or other packages that aren't expecting any novel S3 vectors) from misinterpreting resources, we have `as.character()` return the URI and _not the resource's label_:

```{r}
as.character(food)
```

This can be unexpected. The `table()` function, for example, returns counts by URI:

```{r}
table(kitchen$food)
```

You can use the label by calling `table` on that instead:

```{r}
table(label(kitchen$food))
```

You can convert a linked data frame back into a "normal" data frame (i.e. one not containing vectors of RDF resources) using the `as_dataframe_of_labels()` function. This converts RDF resources into their labels:

```{r}
kitchen_labels <- as_dataframe_of_labels(kitchen)

str(kitchen_labels)
```
